---
title: "Harvard X Capstone Project Movielens"
author: "Motolani Ojo-Bello"
date: "2022-10-29"
output: pdf_document
---
```{r echo=FALSE,warning=FALSE,message=FALSE,cache=TRUE}
#setting download timeout to 100000 seconds
options(timeout=100000)
##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

# Note: this process could take a couple of minutes
# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(xgboost)) install.packages("xgboost", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(xgboost)
library(ggthemes)






# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)


#WRANGLING
#Separating year_of_release from Title column and convert year to numeric
title_year_pattern<- "(\\(\\d\\d\\d\\d\\))"
data_set<- edx%>%mutate(year_of_release=str_extract(title,title_year_pattern),title=str_replace(title,title_year_pattern,""),year_of_release=str_replace(year_of_release,"\\(",""),year_of_release=str_replace(year_of_release,"\\)",""))
data_set<- data_set%>% mutate(year_of_release=as.integer(year_of_release))
#Converting timestamp to datetime format

data_set<- data_set%>%mutate(timestamp=as_datetime(timestamp))
#Using year from timestamp to create new predictor "years_from release"year of review - year of release
data_set<- data_set%>% mutate(year_of_rating=as.integer(year(timestamp)))
##Extract month of rating,day in the month of rating and time of rating in hours
#data_set<-data_set%>%select(!timestamp)
data_set<- data_set%>%mutate(rating_month=as.integer(month(data_set$timestamp)),rating_day=as.integer(day(data_set$timestamp)),nearest_hour_of_rating=as.integer(round(hour(data_set$timestamp)+minute(data_set$timestamp)/60+second(data_set$timestamp)/(60^2))))%>%select(!timestamp)


#Removing movie title since intuitively, it correlates to the movieID
data_set<-data_set%>%select(!title)


#EXPLORATORY ANALYSIS/PLOTS

#plot1 Plotting average rating across genres for top 5 frequently rated genres
to_plot1<-unique(data_set%>%group_by(genres)%>%mutate(n=n())%>% summarise(genres=genres,n=n(),avg=mean(rating)))%>%arrange(desc(n))
plot1<-to_plot1[1:5,]%>%ggplot(aes(genres,avg,group=1))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("Genres")+ylab("Average Rating")+ggtitle("Variation in average rating for the 5 most prevalent Genres")+theme_solarized()
#plot2 Show distribution of ratings for the most frequently rated genre: DRAMA
plot2<-data_set%>%filter(genres%in%to_plot1[1,]$genres)%>% ggplot(aes(rating))+ geom_histogram(binwidth = 1, fill = "antiquewhite4",color="darkgoldenrod4")+xlab("Rating")+ylab("Count")+ggtitle("Distribution of Ratings for the most frequently rated genre")+theme_solarized()

#plot3 Plotting average rating across movie release years for top 5 years with the most ratings 
to_plot3<-unique(data_set%>%group_by(year_of_release)%>%mutate(n=n())%>% summarise(year=year_of_release,n=n(),avg=mean(rating)))%>%arrange(desc(n))
plot3<-to_plot3[1:5,]%>%ggplot(aes(year,avg))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("Year")+ylab("Average Rating")+ggtitle("Variation in average rating for the 5 most prevalent Years of Release")+theme_solarized()
#plot4 Show distribution of ratings for movie release year with most movies: 1995
plot4<-data_set%>%filter(year_of_release%in%to_plot3[1,]$year_of_release)%>% ggplot(aes(rating))+ geom_histogram(binwidth = 1, fill = "antiquewhite4",color="darkgoldenrod4")+xlab("Rating")+ylab("Count")+ggtitle("Distribution of Ratings for movie release year with the most ratings")+theme_solarized()

#plot5 Plotting average rating across year of rating for top 5 years with the most ratings 
to_plot5<-unique(data_set%>%group_by(year_of_rating)%>%mutate(n=n())%>% summarise(year=year_of_rating,n=n(),avg=mean(rating)))%>%arrange(desc(n))
plot5<-to_plot5[1:5,]%>%ggplot(aes(year,avg))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("Year")+ylab("Average Rating")+ggtitle("Variation in average rating for the top 5 prevalent Years of Rating")+theme_solarized()
#plot6 Showing distribution of ratings for year_from_release with the most ratings
plot6<-data_set%>%filter(year_of_rating%in%to_plot5[1,]$year_of_rating)%>% ggplot(aes(rating))+ geom_histogram(binwidth = 1, fill = "antiquewhite4",color="darkgoldenrod4")+xlab("Rating")+ylab("Count")+ggtitle("Distribution of Ratings for the Year with the most ratings")+theme_solarized()


#plot7 Show variation of rating with user ID for the users with the top 5 most ratings 
to_plot7<-unique(data_set%>%group_by(userId)%>%mutate(n=n())%>% summarise(user=userId,n=n(),avg=mean(rating)))%>%arrange(desc(n))
plot7<-to_plot7[1:5,]%>%ggplot(aes(user,avg))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("userId")+ylab("Average Rating")+ggtitle("Variation in average rating for the top 5 prevalent user Id's")+theme_solarized()
#plot8 Show distribution of ratings for user with most movie ratings 
plot8<-data_set%>%filter(userId%in%to_plot7[1,]$userId)%>% ggplot(aes(rating))+ geom_histogram(binwidth = 1, fill = "antiquewhite4",color="darkgoldenrod4")+xlab("Rating")+ylab("Count")+ggtitle("Distribution of Ratings for User with the most ratings")+theme_solarized()

#plot9 Plotting average rating across movieID for top 5 users with the most ratings
to_plot9<-unique(data_set%>%group_by(movieId)%>%mutate(n=n())%>% summarise(movie=movieId,n=n(),avg=mean(rating)))%>%arrange(desc(n))
plot9<-to_plot9[1:5,]%>%ggplot(aes(movie,avg))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("movieId")+ylab("Average Rating")+ggtitle("Variation in average rating for the top 5 prevalent movieId's")+theme_solarized()
#plot10 Showing distribution of ratings for userID with the most ratings
plot10<-data_set%>%filter(movieId%in%to_plot9[1,]$movieId)%>% ggplot(aes(rating))+ geom_histogram(binwidth = 1, fill = "antiquewhite4",color="darkgoldenrod4")+xlab("Rating")+ylab("Count")+ggtitle("Distribution of Ratings for movie with the most ratings")+theme_solarized()

#MORE WRANGLING
#Coverting each possible genre to a predictor rather than having a single Genre predictor with 797 possible values

# Getting list of all possible genres
all_genres<- str_split(data_set$genres,"\\|")
all_genres<-unique(all_genres)
genres<- c()
extractor<- for (i in 1:length(all_genres)) {
  for (j in 1:length(all_genres[[i]])) {
    genres<- c(genres,all_genres[[i]][j])
  }
}
all_genres<- unique(genres) 


#Getting 1 or 0 values for each individually possible genre
Comedy=as.integer(str_detect(data_set$genres,"Comedy")*1)
Romance=as.integer(str_detect(data_set$genres,"Romance")*1)
Action=as.integer(str_detect(data_set$genres,"Action")*1)
Crime=as.integer(str_detect(data_set$genres,"Crime")*1)
Thriller=as.integer(str_detect(data_set$genres,"Thriller")*1)
Drama=as.integer(str_detect(data_set$genres,"Drama")*1)
SciFi=as.integer(str_detect(data_set$genres,"Sci-Fi")*1)
Adventure=as.integer(str_detect(data_set$genres,"Adventure")*1)
Children=as.integer(str_detect(data_set$genres,"Children")*1)
Fantasy=as.integer(str_detect(data_set$genres,"Fantasy")*1)
War=as.integer(str_detect(data_set$genres,"War")*1)
Animation=as.integer(str_detect(data_set$genres,"Animation")*1)
Musical=as.integer(str_detect(data_set$genres,"Musical")*1)
Western=as.integer(str_detect(data_set$genres,"Western")*1)
Mystery=as.integer(str_detect(data_set$genres,"Mystery")*1)
FilmNoir=as.integer(str_detect(data_set$genres,"Film-Noir")*1)
Horror=as.integer(str_detect(data_set$genres,"Horror")*1)
Documentary=as.integer(str_detect(data_set$genres,"Documentary")*1)
IMAX=as.integer(str_detect(data_set$genres,"IMAX")*1)
none=as.integer(str_detect(data_set$genres,"no genres listed")*1)
#combining them into a data frame
genres_df_main<-data.frame(Comedy,Romance,Action,Crime,Thriller,Drama,SciFi,Adventure,Children,Fantasy,War,Animation,Musical,Western,Mystery,FilmNoir,Horror,Documentary,IMAX,none)

#Joining to test set and removing current Genres predictor
data_set<-cbind(data_set,genres_df_main)
data_set<-data_set%>% select(!genres)

#Converting movieID to integer
data_set<-data_set%>%mutate(movieId=as.integer(movieId))


#Applying same wrangling to validation data before running models
validation<- validation%>%mutate(year_of_release=str_extract(title,title_year_pattern),title=str_replace(title,title_year_pattern,""),year_of_release=str_replace(year_of_release,"\\(",""),year_of_release=str_replace(year_of_release,"\\)",""))
validation<- validation%>% mutate(year_of_release=as.integer(year_of_release))
#Converting timestamp to datetime foat

validation<- validation%>%mutate(timestamp=as_datetime(timestamp))
#Using year from timestamp to create new predictor "years_from release"year of review - year of release
validation<- validation%>% mutate(year_of_rating=as.integer(year(timestamp)))

##Extracting month of rating,day in the month of rating and time of rating in hours and removing original timestamp
validation<- validation%>%mutate(rating_month=as.integer(month(validation$timestamp)),rating_day=as.integer(day(validation$timestamp)),nearest_hour_of_rating=as.integer(round(hour(validation$timestamp)+minute(validation$timestamp)/60+second(validation$timestamp)/(60^2))))%>%select(!timestamp)

#removing movie title since it correlates to the movieID
validation<-validation%>%select(!title)

#Getting 1 or 0 values for each individually possible genre
Comedy=as.integer(str_detect(validation$genres,"Comedy")*1)
Romance=as.integer(str_detect(validation$genres,"Romance")*1)
Action=as.integer(str_detect(validation$genres,"Action")*1)
Crime=as.integer(str_detect(validation$genres,"Crime")*1)
Thriller=as.integer(str_detect(validation$genres,"Thriller")*1)
Drama=as.integer(str_detect(validation$genres,"Drama")*1)
SciFi=as.integer(str_detect(validation$genres,"Sci-Fi")*1)
Adventure=as.integer(str_detect(validation$genres,"Adventure")*1)
Children=as.integer(str_detect(validation$genres,"Children")*1)
Fantasy=as.integer(str_detect(validation$genres,"Fantasy")*1)
War=as.integer(str_detect(validation$genres,"War")*1)
Animation=as.integer(str_detect(validation$genres,"Animation")*1)
Musical=as.integer(str_detect(validation$genres,"Musical")*1)
Western=as.integer(str_detect(validation$genres,"Western")*1)
Mystery=as.integer(str_detect(validation$genres,"Mystery")*1)
FilmNoir=as.integer(str_detect(validation$genres,"Film-Noir")*1)
Horror=as.integer(str_detect(validation$genres,"Horror")*1)
Documentary=as.integer(str_detect(validation$genres,"Documentary")*1)
IMAX=as.integer(str_detect(validation$genres,"IMAX")*1)
none=as.integer(str_detect(validation$genres,"no genres listed")*1)
#combining them into a data frame
genres_df_val<-data.frame(Comedy,Romance,Action,Crime,Thriller,Drama,SciFi,Adventure,Children,Fantasy,War,Animation,Musical,Western,Mystery,FilmNoir,Horror,Documentary,IMAX,none)

#Joining to test set and removing current Genres predictor
validation<-cbind(validation,genres_df_val)
validation<-validation%>% select(!genres)

#Converting movieID to integers
validation<-validation%>%mutate(movieId=as.integer(movieId))

#Remove unnecessary variables
rm(genres_df_main,genres_df_val,Action,Adventure,Animation,Children,Comedy,Crime,Documentary,Drama,Fantasy,genres,Horror,IMAX,Musical,Mystery,none,Romance,SciFi,Thriller,War,Western,title_year_pattern,i,j,FilmNoir)



#TABLE SHOWING THAT MOST OF THE GENRES ARE NOT VERY EVENLY DISTRIBUTED
genre_dist<-c(Comedy=mean(data_set$Comedy==1),Romance=mean(data_set$Romance==1),Action=mean(data_set$Action==1),Crime=mean(data_set$Crime==1),Thriller=c(mean(data_set$Thriller==1),Drama=mean(data_set$Drama==1),SciFi=mean(data_set$SciFi==1),Adventure=mean(data_set$Adventure==1),Children=mean(data_set$Children==1),Fantasy=mean(data_set$Fantasy==1),War=mean(data_set$War==1),Animation=mean(data_set$Animation==1),Musical=mean(data_set$Musical==1),Western=mean(data_set$Western==1),Mystery=mean(data_set$Mystery==1),FilmNoir=mean(data_set$FilmNoir==1),Horror=mean(data_set$Horror==1),Documentary=mean(data_set$Documentary==1),IMAX=mean(data_set$IMAX==1),none=mean(data_set$none==1)))
genre_dist<-data_frame(ones=genre_dist)
genre_dist<-genre_dist%>% mutate(zeros=1-ones)
genre_dist<-transpose(genre_dist)
colnames(genre_dist)<-all_genres
genre_dist_plot<-transpose(genre_dist)%>%mutate(genre=all_genres)%>%ggplot(aes(V1,genre,group=1))+geom_point(color="darkgoldenrod4",size=2)+ geom_line(color="deeppink4")+xlab("Proportion of 1's")+ylab("Unique Genres")+ggtitle("Plot showing percentage of 1's in each of the unique Genres")+theme_solarized()
#Creating test and train data
test_index <- createDataPartition(y = data_set$rating, times = 1, p = 0.5,list = FALSE)
test_set<-data_set%>% dplyr::slice(test_index)
train_set<-data_set%>% dplyr::slice(-test_index)


#RMSE Function
RMSE <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
#LM Evaluation
lm_fit<- lm(rating~.,data=train_set)
lm_prediction<-predict(lm_fit,test_set)
lm_RMSE<-RMSE(as.numeric(lm_prediction),test_set$rating)
rm(lm_fit,lm_prediction)

#XGBLINEAR
xg_start_time<-Sys.time()
xg_fit<-train(rating~.,method="xgbLinear",data=train_set,tuneGrid = data.frame(lambda=1e-04, alpha=1e-04, nrounds=150, eta=0.3))
xg_prediction<- predict(xg_fit,test_set)
xg_RMSE<- RMSE(as.numeric(xg_prediction),test_set$rating)
xg_stop_time<-Sys.time()
xg_run_time<-xg_stop_time-xg_start_time
rm(xg_start_time,xg_stop_time)


#Applying XGBOOST FIT TO VALIDATION
final_prediction<- predict(xg_fit,validation)
final_RMSE<- RMSE(as.numeric(final_prediction),validation$rating)
rm(xg_fit)

```



# 1 INTRODUCTION
## INTRODUCTION
The purpose of this project is to build a model that predicts the rating of a movie. The edx data set provided is used to train the model and the final evaluation of the model using the Root Mean Squared Error(RMSE) is carried out on the validation data set. Both data sets(edx and validation) have the same variables:
```{css echo = FALSE}
body {line-height: .8;}
```
- rating: The rating of the movie by a particular user on a scale of 0.5 to 5 at intervals of 5
- user Id: The unique ID of the user who gave the rating
- movie Id: The unique ID of the movie that was rated
- title: The title of the movie accompanied by the year of its release
- time stamp: Unix time stamp of the time the movie was rated 
- genre: A string of all the genres the movie belongs to separated by a '|' 


## KEY STEPS
  The key steps to completing this project included; separating the year of release from the title variable, extracting the year, month, day and nearest hour of rating from the time stamp variable and then separating the genres into variables for each unique genre possible. After this I carried out training and testing of linear regression and xgboost on the edx data set and evaluated their performance using RMSE. Finally, I used the xgboost method on the validation test and evaluated my final RMSE


# 2 VISUALISATION AND METHODOLOGY
## VISUALISATION
For my visualization, I looked at the variation of the average ratings with the predictors genre, year of release, year of rating, user ID and movie ID for the 5 most prevalent values of these predictors and observed general variability based on these predictors as seen in the plots below
```{r ,label="plot batch 1",echo=FALSE, warning=FALSE,message=FALSE,fig.width=11}
plot(plot1)
plot(plot3)
plot(plot5)
plot(plot7)
plot(plot9)
```
I also plotted the distribution of ratings for the most prevalent of each of these predictors and from the plots below, none of the distributions appear normal. This suggests that a linear regression model might not provide the best estimates

```{r ,label="rating distribution plots",echo=FALSE, warning=FALSE,message=FALSE,fig.width=11}
plot(plot2)
plot(plot4)
plot(plot6)
plot(plot8)
plot(plot10)
```


## METHODOLOGY
To start with, using str_extract and str_replace, I removed the year of release in the movie titles and made it a separate predictor year_of_release, after this I removed the title predictor since intuitively it correlates to the movie ID. Then using the mutate and as_datetime functions, I converted the rating time stamps into readable date/time. Following from that, I used the lubridate package to separate the rating date/times into year_of_rating, month_of_rating, day_of_rating and nearest_hour_of_rating. Lastly, looking at the genres predictor it had `r length(unique(edx$genres))` leves, because of this wide variability for a factor, I used str_extract and a nested for loop and created new predictors for all the 20 unique possible genres `r all_genres` and gave them values 1 or 0, 1 if the movie belonged to that genre and 0 otherwise. However, on observing the distribution of these genres in our data set, more than half of the genres are only represented in under 10 percent of our total data as seen in the plot below.This will have an adverse effect on the ability of our model to make accurate predictions
```{r, label="genre distribution plot",echo=FALSE, warning=FALSE,message=FALSE,fig.width=11}
genre_dist_plot
```


# 3 RESULTS
Despite the distributions of the predictors suggesting that a linear regression model might not provide the best predictions I still used that as my baseline model and got an RMSE of `r lm_RMSE` then to improve this I deployed xgboost which uses an ensemble of decision trees to make predictions and got a slightly improved RMSE of `r xg_RMSE`. Finally, applying the xgboost model that was trained with the edx data on the validation set, I obtained a similar RMSE of `r final_RMSE`


# 4 CONCLUSION
Due to the size of our data(`r nrow(data_set)`) and the number of predictors(`r ncol(data_set)-1`) it was difficult to train a more robust model like random forest on our data set with the limitations of a regular computer. Also, a data set with better distribution of genres would likely have yielded a better model. Despite all this, I was still able to estimate ratings on the validation set within an RMSE of `r final_RMSE` which is on average less than one rating point away from the actual rating.
